---
title: "ST588 Project 2 Preparation"
author: "Matt Bray"
format: html
editor: visual
---

# ST588 Preparation/Data Exploration

This code will be used to produce numerical and graphical summaries of the "Store Sales Data" from: [Kaggle](https://www.kaggle.com/datasets/juhi1994/superstore/data). This will allow the developer to get an idea of the data structure and interesting analyses that can be conveyed through interactive tools such as a Shiny app.

Below, we are calling required packages for the code that follows:

```{r}
library(tidyverse)
library(readxl)
library(questionr)
library(vtable)
library(hardhat)
library(terra)
library(geofacet)
library(gganimate)
library(ggpie)
library(shinycssloaders)
```

The following code will import the data from excel into R:

```{r}
data<-read_excel("./US Superstore data.xls")
```

First, we'll do some basic data exploration to understand structure:

```{r}
head(data)
names(data)
```

A couple of variables may be better modelled as factors and we'll rename all variables for programmatic consistency...

```{r}
factors<-c("shipMode", "segment", "country", "state", "postalCode", "region", "category", "subCategory")
data1 <- data |>
  rename("rowID" = "Row ID",
         "orderID" = "Order ID",
         "orderDate" = "Order Date",
         "shipDate" = "Ship Date",
         "shipMode" = "Ship Mode",
         "customerID" = "Customer ID",
         "customerName" = "Customer Name",
         "segment" = "Segment",
         "country" = "Country",
         "city" = "City",
         "state" = "State",
         "postalCode" = "Postal Code",
         "region" = "Region",
         "productID" = "Product ID",
         "category" = "Category",
         "subCategory" = "Sub-Category",
         "productName" = "Product Name",
         "sales" = "Sales",
         "quantity" = "Quantity",
         "discount" = "Discount",
         "profit" = "Profit"
         ) |>
  mutate_at(factors, factor)
```

Now, let's reinspect:

```{r}
head(data1)
names(data1)
str(data1)
```

It appears that the requested variables have been reset as Factors.

## Let's explore some of the categorical variables

First, number of items sold by Segment:

```{r}
#code to see sales by customer (Segment) type:
segFreq <- wtd.table(x=data1[,8], weight=data1[,19])
segFreq
```

Let's Visualize Sales by Segment:

```{r}
ggplot(data = data1, mapping = aes(x=segment, y=quantity, fill=segment)) +
 geom_col() +
 labs(x = "Market Segment",
      y = "Items Sold (#)",
      fill = "Market Segment")
```

Next, number of items sold by State:

```{r}
#code to see sales by state:
stateFreq <- wtd.table(x=data1[,11], weight=data1[,19])
stateFreq
```
Then, let's visualize sales volume by state:
```{r}
data1 |>
  ggplot(data = data1, mapping = aes(x=state, y=quantity)) +
 geom_col() +
 theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
 labs(x = "State",
      y = "Items Sold (#)")
```
Now that we've looked at Frequencies, let's look at some Numerical Summaries.

First, let's see if there if there is anything interesting about sales and profit by Market Segment:
```{r}
ggplot(data = data1, mapping = aes(x=segment, y=sales)) +
 geom_violin() +
 labs(x = "Market Segment",
      y = "Sales")
```


```{r}
ggplot(data = data1, mapping = aes(x=segment, y=profit)) +
 geom_violin() +
 labs(x = "Market Segment",
      y = "Profit")
```

Now, we'll create  number summary to see if perhaps the visualization obscures profitbility, since the boxplots appear centered at 0...

```{r}
segmentProfSum <- sumtable(data=data1, vars="profit", group="segment")
segmentProfSum
```
Are discounted items less profitable?
```{r}
ggplot(data = data1, mapping = aes(x=discount, y=profit, color=year(orderDate))) +
 geom_point() +
 labs(x = "Discount",
      y = "Profit",
      color="Year")
```
It appears that generally, discounted products above 0.2 are not profitble.

What about profit by month?  Is there anything related to the holiday seasons?
```{r}
ggplot(data = data1, mapping = aes(x=month(orderDate, label=TRUE), y=profit)) +
 geom_point() +
 labs(x = "Order Month",
      y = "Profit")
```
Are certain years more profitable?
```{r}
ggplot(data = data1, mapping = aes(x=year(orderDate), y=profit, color=month(orderDate, label=TRUE))) +
 geom_point() +
 labs(x = "Order Year",
      y = "Profit",
      color = "Month")
```


```{r}
    ggplot(data = data1, aes_string(x=data1$segment, y=data1$quantity, fill = data1$segment)) +
      geom_col() +
      coord_flip() +
      facet_geo(~ state) +
      theme_bw() +
      labs(title = paste0("Geographic Representation of quantity by ", data1$segment))
```



